---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
### About

The MLID package provides tools and functions for studying ethnic and social segregation with a multilevel index of dissimilarity that measures uneveness, spatial clustering and scale differences simultaneously.

### Installation

To get the latest version on CRAN:

```{r, eval=FALSE}
installed.packages("MLID")
```

To get the current development version from GitHub:

```{r, eval=FALSE}
# install.packages("devtools")
require(devtools)
devtools::install_github("profrichharris/MLID")
```

### Using MLID

For a quick introduction, follow the examples provided with
```{r, eval=FALSE}
?id
```
For a tutorial
```{r, eval=FALSE}
vignette("MLID", package = "MLID")
```

### Licence

Scripts are released under the [GPL-3 License](https://www.gnu.org/licenses/gpl.html)
